Data Specialization for a Staged Lambda Calculus

### 2015-02-26 paper/intro.tex
minor edits to intro

\newcommand{\langname}{LAMDA12\xspace}
\section{Introduction}

@@

The key difference with partial evaluated program and the stratified
program is that the latter has lost no generality: it works with all
arguments, whereas the partially evaluated program is specialized for
a particular value of $x$ ($x = [~]$ in the example).  Thus, the pass
separation can be applied without leading to any loss of generality.
In fact, this is quite natural: going back to the terminology of
Jorring and Scherlis, pass separation is a frequency reduction,
whereas as partial evaluation and meta-programming are precomputation
techniques. Consequently, if desired, partial evaluation can be
orthogonally applied to a stratified program.

@@

But what is the benefit or pass separation? The big benefit is
efficiency. By stratifying the program, pass separation factors out
frequently used the parts of computation, computes them first, and
then reuse them later instead of computing them again.  Since it can
be applied statically with no loss of generality, pass separation can
significantly improve performance.  In fact, in computer graphics,
program are typically manually stratified both for improved
efficiency. Such stratification is in fact required by the execution
model of some modern graphics architectures.

@@

In this paper we make significant progress on the problem of automatic
stratification of higher-order functional programs.  As our starting
point, we take the modal language \langname with explicit staging in
the style of Davies~\cite{davies96}, where the ``circle'' modality
denotes computation in a later stage, and present a staged operational
semantics similar for \lang\ (\ref{sec:stagedsemantics}).  We then
present a splitting algorithm that stratifies a \langname program into
a two mono-staged programs, one for the first stage and for the second
stage (\ref{sec:splitting}).  Since each program is monostaged, they
are expressed in conventional functional languages.

>>>>

\section{Introduction}

@@

The key difference with partial evaluated program and the stratified
program is that the latter has lost no generality: it works with all
arguments, whereas the partially evaluated program is specialized for
a particular value of $x$ ($x = [~]$ in the example).  Thus, the pass
separation can be applied without leading to any loss of generality.
In fact, this is quite natural---in the terminology of
J{\o}rring and Scherlis, pass separation is a frequency reduction,
whereas as partial evaluation and metaprogramming are precomputation
techniques. Consequently, if desired, partial evaluation can be
orthogonally applied to a stratified program.

@@

But what is the benefit of pass separation? The big benefit is
efficiency. By stratifying the program, pass separation factors out
frequently used the parts of computation, computes them first, and
then reuse them later instead of computing them again.  Since it can
be applied statically with no loss of generality, pass separation can
significantly improve performance.  In fact, in computer graphics,
program are typically manually stratified both for improved
efficiency. Such stratification is in fact required by the execution
model of some modern graphics architectures.

@@

In this paper we make significant progress on the problem of automatic
stratification of higher-order functional programs.  As our starting
point, we take the modal language \lang\ with explicit staging in
the style of Davies~\cite{davies96}, where the ``circle'' modality
denotes computation in a later stage, and present a staged operational
semantics similar for \lang\ (\ref{sec:stagedsemantics}).  We then
present a splitting algorithm that stratifies a \lang\ program into
a two mono-staged programs, one for the first stage and for the second
stage (\ref{sec:splitting}).  Since each program is monostaged, they
are expressed in conventional functional languages.
